Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-10-17 paper/sections/conclusions.tex
[ paper ] conclusions reviewed.

We presented a reliable proof-reconstruction of \Metis solutions
for problems in classical propositional logic without logic.
The reconstruction approach in this study was designed to use
only a syntactic treatment of the logic without considering the
semantics. This aspect represent a difficulty but also an opportunity
to extend this work towards the support for first-order logic.
As a first future work, we see the support of \Metis propositional
fragment with equality.

Summarizing, we provided for each \Metis inference rule a theorem that
emulate the rule deduction. The emulated
version rules exposed in Section~\ref{ssec:emulating-inferences} admit
improvements, we left this for future work, particularly, a better
implementation of \simplify rule.
Some of the improvements are related with computational efficiency.
For instance, we use widely functions to check for syntactical equivalence
between propositions and to reorder propositions to match with another.
We believe both are expensive tasks and can get help in the analysis
step inside the translator tool.

The proof-reconstruction tool presented in this paper
is able to translate \Metis derivations in \TSTP format to \Agda proof-terms.
The development presented in this paper can be extended to support
other \ATPs for \CPL like \name{EProver} or \name{Z3}.
This future work can be carried out by following the \name{EProver} inference
rules described on Kanso' Ph.D. Thesis~\cite{Kanso2012}.

One of the main contribution of this study is
increasing the trustworthiness of these automatic provers,
in our case, \Metis. Justify a proof-object of a theorem prover has a real
significant impact for these automatic tools.
The reverse engineering task can reveal important issues or bugs
in many parts of these system (e.g., preprocessing, reasoning, or
deduction modules). While doing this research, we had the opportunity
to find some of those kind of bugs\footnote{Issues No.~2, No.~4. and
commit \name{8a3f11e} in the \Metis' official repository
\url{https://github.com/gilith/metis}}. Fortunately, all these bugs
were fixed quickly for release 20170810 by Joe Hurd.

>>>>

We presented a reliable proof-reconstruction in type theory of \Metis
solutions for problems in classical propositional logic without
equality. As a product of this research, we built a proof-reconstruction tool written in \Haskell. This tool generates
\Agda proof-terms of \Metis \TSTP derivations. \Agda files
delivered by the aforementioned tool are supported by the
formalization in \Agda of the \Metis rules described in
Section~\ref{ssec:emulating-inferences}.

The reconstruction approach in this study was designed to use
only syntactical aspects of the logic. This decision was first a
difficulty since it demands more detailed proof-objects, a
description of every transformation or deduction step performed by
the prover, see
Section~\ref{sssec:normalization} and Section~\ref{sssec:reordering}.
But, we chose that syntactical treatment as the opportunity
to extend this work towards the support of first-order logic.
Recall, satisfiability in first-order logic is undecidable.

As an early future work, we propose to extend the
proof-reconstruction presented in this paper to support the
proposition fragment with equality of \Metis.
Another option is to extend the proof-reconstruction to support other \ATPs for \CPL like \name{EProver} or \name{Z3}.
This development can be carried out by following the \name{EProver}
inference rules described on Kanso' Ph.D. Thesis~\cite{Kanso2012}.

Summarizing, we provided for each \Metis inference rule a theorem in
type theory to emulate the rule reasoning. This approach mainly
exposed in Section~\ref{ssec:emulating-inferences}
admits some improvements, we leave it for future work. We concern
mainly for a better coverage of the \simplify rule.
Other improvements are related with computational efficiency of
rules strongly depend on reordering tasks,
Section~\ref{sssec:reordering}.
We believe some rules like \canonicalize can get some help from a analysis done in the translator tool.

One of the main contribution of this study was
increasing the trustworthiness of the automatic provers,
in our case, \Metis.
Justifying a proof-object of a theorem prover
has a real significant impact for these automatic tools.
The reverse engineering task can reveal important issues or bugs
in many parts of these system (e.g., preprocessing, reasoning, or
deduction modules). During this research, we had the opportunity
to find some of bugs\footnote{Issues No.~2, No.~4. and
commit \name{8a3f11e} in the \Metis' official repository
\url{https://github.com/gilith/metis}} in \Metis.
Fortunately, all these problems were fixed quickly by Joe Hurd in
20170810 release.

### 2017-10-17 paper/sections/proof-reconstruction.tex
[ paper ] conclusions reviewed.

\subsubsection{Reordering.}
Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of the propositions, our approach shows equivalence
by converting propositions to their conjunctive normal forms.
But also, reordering the conjunctions and the inner disjunctions into the
formulas to match them.
In this section, we provide some theorems to perform such reordering tasks,
the goals are the following.

@@

\subsubsection{Normalization.}
\label{sssec:normal_forms}
In the core of \Metis, we find an important module:
\texttt{Normalize.sml}.
This module is the responsible for rules like \canonicalize, \clausify,
\conjunct or \simplify.
One of the main purposes of the module consists of adding the axioms to the problem. Into this process, some transformation are carried out by different inference rules.
For example, the \canonicalize rule can convert a first-order form formulas to their clausal normal form. In that process, the formula can suffers different transformations by removing redundancies (tautologies or definitions), converting the proposition to its normalized negative normal form, to its conjunctive normal form, or to a specialized normal form.

@@

A sequent $Γ ⟝ φ$ represents a theorem where
$Γ$ is a set of \Prop propositions premises, and $φ : \NProp$ is the
sequent's conclusion. These theorems have a subset of rules from
the classical inference rules in Fig.~\ref{fig:CPL-inference-rules}
that apply for the connectives that coincides with \NProp type.

@@

In a disjunction, $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$, we say $ψ ∈_{∨} φ$,
if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.
Note that $ψ ∈_{∨} φ$ can be defined as the equality
$ψ ≡ conjunct_{∨}(φ, ψ)$.

@@

Now, we have removed redundancies in the disjunctions by applying
$rm_{∨}$ function, we can define a similar function to work with conjunctions.

@@

We are ready to combine all function defined above in two function
$canon_{∧}$ for removing redundancies in conjunctions, and
$canon_{∨}$ for removing redundancies in disjunctions.

@@

Finally, we can translate formulas from \Prop to \NProp by defining the
$nnf₀$ function based on the \Metis'source code. We have taken
inspiration on the $nnf$ function described in
\citeauthor{Bezem2002}~\cite{Bezem2002}. Their $nnf$ function translates
formulas to their classical negative normal form. In the reference, the
authors avoid a termination problem with a similar function by using as an
extra argument, the polarity of the formula.

For the polarity of a formula, we use the \abbre{Polarity} type with two
constructors: $⊕$ to denote positive polarity, and $⊖$ to denote negative
polarity.

@@

\begin{sketchproof} If $φ ≡ φ₁ ∨ φ₂$, we get for $i = 1, 2$,
\begin{equation*}
  \begin{bprooftree}
    \RightLabel{assume}
    \AxiomC{$Γ, φ_{i} ⊢ φ_{i}$}
    \RightLabel{lem-dist}
    \UnaryInfC{$Γ ⊢ dist(φ_{i})$}
    \RightLabel{∨-intro$_{1,2}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist(φ₁) ∨ dist(φ₂)$}
    \RightLabel{lem-dist$_{∨}$.}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$}
    \end{bprooftree}
\end{equation*}
Then, by using the ∨-elim rule, from the last derivation above we derive
$Γ, φ₁ ∨ φ₂ ⟝ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$, and the lemma follows.
\end{sketchproof}

@@

\begin{theorem}[thm-cnf]
\label{thm:thm-cnf}
  $Γ ⊢ φ$ then $Γ ⊢ cnf(φ)$, where,
  \begin{align*}
    \begin{split}
    &cnf : \NProp \to \NProp\\
    &cnf = form ∘ dist ∘ nnf.
    \end{split}
  \end{align*}
\end{theorem}

@@

Since all transformations performed above come from equivalences
in classical propositional logic, we can also states the following theorem
convenient for the next section.

@@

The \canonicalize rule is an overloaded inference rule that performs
mainly normalization of the proposition input by using theorems
described in the section above.  It can introduce axioms and definitions
in the problem. This rule convert a \texttt{fof} %TODO
formula in clausal form to a \CNF clause but also it perform some
simplification in the process (removing ⊤ and ⊥ from the formula).

@@

\begin{proof} When $φ ≡ ψ$, we use the $subs$ theorem Eq.~(\ref{eq:substitution-theorem}). Otherwise, we use Theorem \ref{thm:thm-reorder-and-or} and Theorem \ref{thm:thm-cnf}.
\end{proof}

>>>>

\subsubsection{Reordering.}
\label{sssec:reordering}

Logic equivalence between propositions is a major issue to justify
prover reasoning steps. Since we left out semantics to treat only the
syntax aspects of the propositions, our approach shows equivalence
by converting propositions to their conjunctive normal forms.
But also, reordering the conjunctions and the inner disjunctions into the
formulas to match them.
In this section, we provide some theorems to perform such reordering tasks,
the goals are the following.

@@

\subsubsection{Normalization.}
\label{sssec:normalization}
The normalization module in \Metis is the responsible for
\canonicalize, \clausify, \conjunct and \simplify.
One of the main purposes of the module consists of adding the axioms
to the problem. In this process, some transformations are carried
out by convert a \verb!fof! formula to their clausal normal form.
The process mainly consists of removing redundancies (tautologies or
definitions) and converting the proposition to its normalized
negative normal form, conjunctive normal form, or
definitional normal form.

@@

A sequent $Γ ⟝ φ$ represents a theorem where
$Γ$ is a set of \Prop propositions premises, and $φ : \NProp$ is the
conclusion of the theorem. These theorems have a subset of rules from
the classical inference rules in Fig.~\ref{fig:CPL-inference-rules}
that apply for the connectives that coincides with \NProp type.

@@

In a disjunction, $φ = φ₁ ∨ φ₂ ∨ \cdots ∨ φₙ$, we say $ψ ∈_{∨} φ$,
if there is some $i = 1, \cdots, n$ such that $ψ = φᵢ$.
$ψ ∈_{∨} φ$ is another name for the equality
$ψ ≡ conjunct_{∨}(φ, ψ)$.

@@

Now, we have removed redundancies in the disjunctions by applying
$rm_{∨}$ function, we can define a similar function to work with
conjunctions, the $rm_{∧}$ function.

@@

We use the functions defined above to remove redundancies in conjunctions with $canon_{∧}$ function, and
$canon_{∨}$ function to remove redundancies in disjunctions.

@@

Finally, we can translate formulas from \Prop to \NProp by defining
the $nnf₀$ function. We have taken
inspiration on the $nnf$ function described in
\citeauthor{Bezem2002}~\cite{Bezem2002}. Their $nnf$ function
translates formulas to their classical negative normal form. In the
reference, the authors avoid a termination problem with a similar
function by using as an extra argument, the polarity of the formula.
We follow the same idea in Def.~\ref{def:nnf-zero}.
Therefore, we need the \abbre{Polarity} type with
two constructors: $⊕$ to denote positive polarity, and $⊖$ to denote
negative polarity of the formula.

@@

\begin{sketchproof} If $φ ≡ φ₁ ∨ φ₂$, we get for $i = 1, 2$,
\begin{equation*}
  \begin{bprooftree}
    \RightLabel{assume}
    \AxiomC{$Γ, φ_{i} ⊢ φ_{i}$}
    \RightLabel{lem-dist}
    \UnaryInfC{$Γ ⊢ dist(φ_{i})$}
    \RightLabel{∨-intro$_{1,2}$}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist(φ₁) ∨ dist(φ₂)$}
    \RightLabel{lem-dist$_{∨}$.}
    \UnaryInfC{$Γ, φ_{i} ⊢ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$}
    \end{bprooftree}
\end{equation*}
Then, by using the ∨-elim rule in the last derivation above, we
derive $Γ, φ₁ ∨ φ₂ ⟝ dist_{∨}(dist(φ₁) ∨ dist(φ₂))$ and the lemma
follows.
\end{sketchproof}

To convert normalized propositions to classical propositions, we
make the following definition, completing the translation between
these two types.

@@

As a result, we can get a conjunctive normal form by applying the
following theorem.

\begin{theorem}[thm-cnf]
\label{thm:thm-cnf}
  $Γ ⊢ φ$ then $Γ ⊢ cnf(φ)$, where,
  \begin{align*}
    \begin{split}
    &cnf : \Prop \to \Prop\\
    &cnf = form ∘ dist ∘ nnf.
    \end{split}
  \end{align*}
\end{theorem}

@@

Since all transformations performed above come from equivalences
in classical propositional logic, we have stated the following theorem with a purpose for the next section.

@@

The \canonicalize rule is an overloaded inference rule that performs
mainly normalization of the proposition input by using theorems
described in the section above.  It can introduce axioms and
definitions into the problem solution.
This rule convert a \texttt{fof} %TODO
formula in clausal form to a \CNF clause but also it perform some
simplification in the process (e.g., removing ⊤ and ⊥ in the formula).

@@

\begin{proof} If $φ ≡ ψ$, this lemmas follows after applies $subs$
lemma. Otherwise, we use Theorem~\ref{thm:thm-reorder-and-or} and Theorem~\ref{thm:thm-cnf}.
\end{proof}

### 2017-10-17 paper/sections/related-work.tex
[ paper ] conclusions reviewed.

Kanso and Setzer~\cite{kanso2016light} integrate the propositional fragment of \name{EProver} in \Agda. In this work, the authors translate automatically propositional logic problems written in \Agda to \TPTP problems. By calling an external wrapper the ATP, this tool gets a \TSTP derivation. The tool parses the \name{EProver} derivation and outputs
\Agda code. After that, we say their proof-reconstruction is almost
similar as the approach presented in Section~\ref{ssec:workflow}.
Nevertheless, Kanso and Setzer aim to justify some of the inference rules
by using propositional semantics instead of just syntactic derivation
steps.
\end{document}

>>>>

Kanso and Setzer~\cite{kanso2016light} integrate the propositional fragment of \name{EProver} in \Agda. In this work, the authors translate automatically propositional logic problems written in \Agda to \TPTP problems. By calling an external wrapper of the ATP, this tool gets a \TSTP derivation. The tool parses the \name{EProver} derivation and outputs
\Agda code. After that, we say their proof-reconstruction is almost
similar as the approach presented in Section~\ref{ssec:workflow}.
Nevertheless, Kanso and Setzer aim to justify some of the inference rules
by using propositional semantics instead of a only syntactic treatment. We showed here a kind of treatment to justify \Metis proof-objects.
\end{document}
